taxadb: A High-Performance Local Taxonomic Database Interface

### 2020-01-23 README.Rmd
Add code of conduct
Change URL after ropensci migration

[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![Travis build status](https://travis-ci.org/cboettig/taxadb.svg?branch=master)](https://travis-ci.org/cboettig/taxadb)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/cboettig/taxadb?branch=master&svg=true)](https://ci.appveyor.com/project/cboettig/taxadb)
[![Coverage status](https://codecov.io/gh/cboettig/taxadb/branch/master/graph/badge.svg)](https://codecov.io/github/cboettig/taxadb?branch=master)
[![CRAN status](https://www.r-pkg.org/badges/version/taxadb)](https://cran.r-project.org/package=taxadb)

@@

```{r eval=FALSE}
devtools::install_github("cboettig/taxadb")
```

@@

>>>>

[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![Travis build status](https://travis-ci.org/ropensci/taxadb.svg?branch=master)](https://travis-ci.org/ropensci/taxadb)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/ropensci/taxadb?branch=master&svg=true)](https://ci.appveyor.com/project/ropensci/taxadb)
[![Coverage status](https://codecov.io/gh/ropensci/taxadb/branch/master/graph/badge.svg)](https://codecov.io/github/ropensci/taxadb?branch=master)
[![CRAN status](https://www.r-pkg.org/badges/version/taxadb)](https://cran.r-project.org/package=taxadb)

@@

```{r eval=FALSE}
devtools::install_github("ropensci/taxadb")
```

@@

----

Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md).
By participating in this project you agree to abide by its terms.

[![ropensci_footer](https://ropensci.org/public_images/ropensci_footer.png)](https://ropensci.org)

### 2020-01-23 paper/manuscript.Rmd
Change URL after ropensci migration

```{r message=FALSE, include = FALSE}
library(kableExtra)
library(magrittr)
library(tidyverse)
library(taxadb)
library(printr)
library(knitr)

@@

printtable <- function(df){
  df %>%
    kableExtra::kable("latex", booktabs=T) %>%
    kableExtra::kable_styling(full_width = F)
}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE)

@@

As ecologists and evolutionary biologists synthesize datasets across larger and larger assemblies of species, we face a continual challenge of reconciling taxonomic names. How many species are in the combined data? Do the studies use the same names for the same species, or do they use different synonyms for the same species?
Failing to correct for such differences can lead to significant inflation of species counts and miss-aligned  datasets (Figure 1).
These challenges have become particularly acute as it becomes increasingly common for researchers to work across larger number and diversity of species in any given analysis, which may preclude the resources or substantative taxonomic expertise all clades needed to resolve scientific names [@Patterson2010].
While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008, @maldonado2015], and a growing number of databases and tools have emerged and grown over the past few decades [e.g. @itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018], it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
Here, we present an R package, `taxadb`, which seeks to address this gap.

@@

```{r eval=FALSE}
remotes::install_github("cboettig/taxadb")
```
-->

@@

By default `taxadb` creates a MonetDBLite database instance [@monetdblite], a columnar-oriented relational database requiring no additional installation while also providing persistent disk-based storage. `taxadb` will automatically handle opening, caching, and closing the database connection.

@@

`get_ids` offers a first pass at matching scientifc names to id, but names may remain unresolved for a number of reasons. First, a name may match to multiple accepted names, as in the case of a species that has been split. By design, these cases are left to be resolved by the researcher using the 'filter_' functions to filter underlying taxonomic tables for additional information. A name may also be unresolved due to typos or improper formatting. `clean_names` addresses common formatting issues such as the inclusion of missing species epithets (e.g. `Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithets such as `Colaptes auratus cafer` that prevent matches to the bionmial name. These modifications are not appropriate in all settings and should be used with care (Language from vignette). Spell check of input names is outside the scope of `taxadb`.

Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym. Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@Franz2009, @franz2018]. The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".

@@

```{r echo = FALSE, results = "asis"}
#provider descriptions
desc <- c(itis = "originally formed to standardize taxonomic name usage across many agencies in the United States federal government",
          ncbi = "nomenclature for sequences in the International Nucleotide Sequence Database Collaboration database",
          col = "comphrehensive taxonomic effort, includes some other providers (e.g. itis)",
          gbif = "taxonomic backbone of the GBIF database, assembled from other sources including COL",
          fb = "nomenclature for global database of fishes",
          ott = "comphrenesive tree of life based on phylogenetic trees and taxonomic data",
          iucn = "taxonomy for classification of species status")

tibble(provider = c("Integrated Taxonomic Information System (ITIS 2019)",
                                 "National Center for Biological Information's Taxonomy database (NCBI)",
                                 "Catalogue of Life (COL)",
                                 "Global Biodiversity Information Facility Taxonomic Backbone (GBIF 2019)",
                                 "FishBase (Froese and Pauly 2019)",
                                 "Open Tree Taxonomy (OTT)",
                                 "International Union for Conservation of Nature and Natural Resources (IUCN 2019)"),
                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", "ott", "iucn"),
                    total_identifiers = map(abbreviation,
                                            ~taxa_tbl(.x) %>%
                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct()),
                    description = desc
) %>%
  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE) %>%
  kable_styling(latex_options="scale_down") %>%
  kableExtra::column_spec(1, width = "5cm") %>%
  kableExtra::column_spec(4, width = "10cm")

@@

Taxonomic indentifiers are a powerful tool for maintaing taxonomic consistency, an essential task for a wide variety of applications. Despite multiple taxonomic standardization efforts, resolving names to taxonomic identifiers is often not a standard step in the research work flow due to difficulty in accessing providers and the time consuming API queries necessary for resolving even moderately sized data sets. `taxadb` fills an important gap between existing tools and typical research patterns by providing a fast, reproducible approach for matching names to taxonomic identifiers. `taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution. In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.

>>>>

```{r message=FALSE, include = FALSE}
library(kableExtra)
library(magrittr)
library(tidyverse)
library(taxadb)
library(printr)

@@

printtable <- function(df){
  df %>% 
    kableExtra::kable("latex", booktabs=T) %>% 
    kable_styling(full_width = F)
}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE)

@@

As ecologists and evolutionary biologists synthesize datasets across larger and larger assemblies of species, we face a continual challenge of reconciling taxonomic names.
How many species are in the combined data? Do the studies use the same names for the same species, or do they use different synonyms for the same species?
Failing to correct for such differences can lead to significant inflation of species counts and miss-aligned  datasets (Figure 1).
These challenges have become particularly acute as it becomes increasingly common for researchers to work across larger number and diversity of species in any given analysis, which may preclude the resources or substantative taxonomic expertise all clades needed to resolve scientific names [@Patterson2010].
While these issues have long been recognized in the literature [], and a growing number of databases and tools have emerged and grown over the past few decades [e.g. @itis; @ncbi; @col], it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
Here, we present an R package, `taxadb`, which seeks to address this gap.

@@

```{r eval=FALSE}
remotes::install_github("ropensci/taxadb")
```
-->

@@

<!--
`taxadb` does not require the use of an external database system, but will automatically make use of the fastest database server available. `taxadb` can also use any DBI compliant database accessible from R, such as MySQL or Postgres [see @DBI], though the default `duckdb` or `MonetDBLite` will be both significantly faster and easier to use.  
-->

<!--  Next 2 paragraphs should possibly be left to a manual, but are likely to be overlooked aspects there... -->
By default `taxadb` creates a MonetDBLite database instance [@monetdblite], a columnar-oriented relational database requiring no additional installation while also providing persistent disk-based storage. `taxadb` will automatically handle opening, caching, and closing the database connection.

<!-- 
now sidestepped

but note that one limitation of MonetDBLite is the restriction against concurrent access: two separate R sessions on the same machine cannot both access the database at the same time. Users who need concurrent access from multiple sessions may wish to use a standalone MonetDB server or other database server instead.
-->

@@

`get_ids` offers a first pass at matching scientifc names to id, but names may remain unresolved for a number of reasons. First, a name may match to multiple accepted names, as in the case of a species that has been split. By design, these cases are left to be resolved by the researcher using the 'filter_' functions to filter underlying taxonomic tables for additional information. A name may also be unresolved due to typos or improper formatting. `clean_names` addresses common formatting issues such as the inclusion of missing species epithets (e.g. `Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithets such as `Colaptes auratus cafer` that prevent matches to the bionmial name. These modifications are not appropriate in all settings and should be used with care (Language from vignette).

Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym. Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider. The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".

@@

Table version of below - citation, number of unique identifier

- the Integrated Taxonomic Information System [ITIS; @itis], original formed to standardize taxonomic name usage across many agencies in the United States federal government, 
- the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi]
- Catalogue of Life (COL) Annual Species list. (2018 version at present)
- Global Biodiversity Information Facility Taxonomic Backbone [GBIF; @gbif].  
- FishBase [@fishbase] taxonomic names
- SeaLifeBase [@sealifebase] taxonomic names
- Wikidata [@jorrit]
- Open Tree Taxonomy, OTT
- International Union for Conservation of Nature and Natural Resources (IUCN) Red list @iucn

<!--
What other aspects should be included in the table? Citations?
-->

```{r cache = FALSE, echo = FALSE}
tibble(provider = c("Integrated Taxonomic Information System", 
                                 "National Center for Biological Information's Taxonomy database",
                                 "Catalogue of Life",
                                 "Global Biodiversity Information Facility Taxonomic Backbone",
                                 "FishBase",
                                 #"SeaLifeBase",
                                 "Wikidata", "Open Tree Taxonomy",
                                 "International Union for Conservation of Nature and Natural Resources"),
                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", #"slb",
                                     "wd", "ott", "iucn"),
                    total_identifiers = map(abbreviation, 
                                            ~taxa_tbl(.x) %>% 
                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct())
) %>% printtable()

@@

Taxonomic indentifiers are a powerful tool for maintaing taxonomic consistency, an essential task for a wide variety of applications. Despite multiple taxonomic standardization efforts, resolving names to taxonomic identifiers is often not a standard step in the research work flow due to difficulty in accessing providers and the time consuming API queries necessary fo resolving even moderately sized data sets. `taxadb` fills an important gap between existing tools and typical research patterns by providing a fast, reproducible approach for matching names to taxonomic identifiers. `taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution.In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.

`taxadb` establishes a local database of provider allowing for consistent and reproducible queries. The set of functions operate consistently accross providers returning data in a standardized format.

----
